Untwisting the Boerdijk-Coxeter Helix

### 2017-05-26 doc/TetrahelixUntwist.tex
teaks

$\theta_{bc}$ is approximately $0.37 \cdot 2 \pi$ radians or  $131.81$ degrees.
The angle $\theta_{bc}$ is the rotation of \emph{each} tetrahedron, not the tetraheda along a rail.  In \cref{fig:closeup},
each tetrahedron has either a yellow, blue, or red outer edge or rail.
That is, a blue-rail tetrahedron is rotated slightly more than a $1/3$ of a revolution to match the face of the yellow tetrahedra.

R.W. Gray's website\cite{graytetrahelix}, repeating a formula by Coxeter\cite{coxeter1985simplicial} in a more accessible form, gives the Cartesian coordinates $\begin{bmatrix}
           x \\
           y \\
           z
         \end{bmatrix}$
for a counter-clockwise BC Helix in a right-handed coordinate system:
\begin{equation}
  \label{graycoxeter}
\vec{V}(n) =
\left [
  \begin{tabular}{c}
   $ r_{bc} \cos{n \theta_{bc}} $\\
   $ r_{bc} \sin{n \theta_{bc}} $\\
   $ n h_{bc}  $
  \end{tabular}
  \right ],
\text{where:}
  \begin{tabular}{c}
 $ r_{bc} = \frac{3\sqrt{3}}{10} \approx 0.5196 $\\
 $ h_{bc} = 1/\sqrt{10} \approx 0.3162 $ \\
 $ \theta_{bc} = \arccos(-2/3) $ \\
  \end{tabular}      
\end{equation}
where $n$ represents each integer numbered node in succession on every colored rail.

@@

The $\vec{H}_{BCcolored}(n,c)$ formulation can be further clarified by rewriting directly in terms of the rail angle $\rho_{bc}$ rather than $\theta_{bc}$.
Intuitively we seek an expression where $c/3$ is multiplied by a $1/3$ rotation plus the rail angle $\rho$.
We expand 
the expressions $\theta_{bc}$ and $\rho_{bc}$ in \eqref{eq:colored} and seek to isolate the term $c 2\pi/3 $.
\begin{align*}
 c \theta_{bc}  &=   \text{\{we aim for 3 in denominator, so we split...\}} \\
    (c/3)  (3 \theta_{bc})  &=   \text{\{we want $2\pi$ in numerator, so add canceling terms...\}} \\
 (c/ 3) ((3 \theta_{bc} - 2 \pi)  + 2 \pi) &= \text{\{definition of $\rho_{bc}$\}...} \\  
  (c / 3) \rho_{bc}  + c 2 \pi /3 &=  \text{\{algebra...\}} \\  
c  ( \rho_{bc} +  2 \pi) /3  \\
\end{align*}

@@

In the proof below we find useful to consider projection diagrams that are the axial projection
of a tetrahelix onto the $XY$-plane. \Cref{fig:projectiondiagram} is an example of such a diagram.
\begin{lemma}
  If the rail angle $0 < \rho < \pi$ is a rational multiple of $\pi$, then the projection of
  edges long a helix of that rail angle  along the $z$-axis onto the $XY$-plane form a
  regular polygon of 3 or more sides, else they fill in a complete circle.
  \label{lemma:ngon}
\end{lemma}
\begin{proof}
  All points lying on a helix projected along the axis lie on a circle in the $XY$-plane.
  Helices are periodic in the $z$ dimension modulo $2\pi$.
  If $2\pi/\rho$ is irrational, the projection onto the $XY$-plane will contain
  an unbounded number of points on a circle.
  If and only if $2\pi/\rho$ is rational, the projection onto
  the $XY$-plane will contain a finite number of points.
  Because $\pi$ is transcendental and irrational,
  $2\pi/\rho$ is rational if and only $\rho = a\pi/b$, where $a$ and $b$ are integers
  and without loss of generality $a$ and $b$ are coprime.
  $\rho < \pi$, therefore $a < b$. Also, $\rho > 0$, therefore $a > 0$.
  The number of points in the projection is $2b$ if $a$ is odd, and $b$ if $a$ is even.
  This polygon has at least 3 sides, since either $\rho$ is irrational or $b > a$, and therefore
  $b >= 2$. If $a/b = 1/2$, the projection is a square, which has four sides. 
\end{proof}

>>>>

Note that $\theta_{bc}$ is approximately $0.37 \cdot 2 \pi$ radians or  $131.81$ degrees.
The angle $\theta_{bc}$ is the rotation of \emph{each} tetrahedron, not the tetraheda along a rail.  In \cref{fig:closeup},
each tetrahedron has either a yellow, blue, or red outer edge or rail.
That is, a blue-rail tetrahedron is rotated slightly more than a $1/3$ of a revolution to match the face of the yellow tetrahedra.

R.W. Gray's website\cite{graytetrahelix}, repeating a formula by Coxeter\cite{coxeter1985simplicial} in a more accessible form, gives the Cartesian coordinates $\begin{bmatrix}
           x \\
           y \\
           z
         \end{bmatrix}$
for a counter-clockwise BC Helix in a right-handed coordinate system:
\begin{equation}
  \label{graycoxeter}
\vec{V}(n) =
\left [
  \begin{tabular}{c}
   $ r_{bc} \cos{n \theta_{bc}} $\\
   $ r_{bc} \sin{n \theta_{bc}} $\\
   $ n h_{bc}  $
  \end{tabular}
  \right ],
\text{where:}
  \begin{tabular}{c}
 $ r_{bc} = \frac{3\sqrt{3}}{10} \approx 0.5196 $\\
 $ h_{bc} = 1/\sqrt{10} \approx 0.3162 $ \\
 $ \theta_{bc} = \arccos(-2/3) $ \\
  \end{tabular}      
\end{equation}
and where $n$ represents each integer numbered node in succession on every colored rail.

@@

The $\vec{H}_{BCcolored}(n,c)$ formulation can be further clarified by rewriting directly in terms of the rail angle $\rho_{bc}$ rather than $\theta_{bc}$.
Intuitively we seek an expression where $c/3$ is multiplied by a $1/3$ rotation plus the rail angle $\rho$.
We expand 
the expressions $\theta_{bc}$ and $\rho_{bc}$ in \eqref{eq:colored} and seek to isolate the term $c 2\pi/3 $.
\begin{align*}
 c \theta_{bc}  &=   \text{\{we aim for 3 in denominator, so we split...\}} \\
    (c/3)  (3 \theta_{bc})  &=   \text{\{we want $2\pi$ in numerator, so add canceling terms...\}} \\
 (c/ 3) ((3 \theta_{bc} - 2 \pi)  + 2 \pi) &= \text{\{definition of $\rho_{bc}$\ ...\}} \\  
  (c / 3) \rho_{bc}  + c 2 \pi /3 &=  \text{\{algebra...\}} \\  
c  ( \rho_{bc} +  2 \pi) /3  \\
\end{align*}

@@

In the proof below we find useful to consider projection diagrams that are the axial projection
of a tetrahelix onto the $XY$-plane. \Cref{fig:projectiondiagram} is an example of such a diagram.
\begin{lemma}
  If the rail angle $0 < \rho < \pi$ is a rational multiple of $\pi$, then the projection of
  edges long a helix of that rail angle  along the $z$-axis onto the $XY$-plane form a
  regular polygon of 3 or more sides, else they fill in a complete circle.
  \label{lemma:ngon}
\end{lemma}
\begin{proof}
  All points lying on a helix projected along the axis lie on a circle in the $XY$-plane.
  Helices are periodic in the $z$ dimension modulo $2\pi$.
  If $2\pi/\rho$ is irrational, the projection onto the $XY$-plane will contain
  an unbounded number of points on a circle.
  If and only if $2\pi/\rho$ is rational, the projection onto
  the $XY$-plane will contain a finite number of points.
  Because $\pi$ is transcendental and irrational,
  $2\pi/\rho$ is rational if and only $\rho = a\pi/b$, where $a$ and $b$ are integers
  and without loss of generality $a$ and $b$ are coprime.
  Since $\rho < \pi$, therefore $a < b$. Also, $\rho > 0$, therefore $a > 0$.
  The number of points in the projection is $2b$ if $a$ is odd, and $b$ if $a$ is even.
  This polygon has at least 3 sides, since either $\rho$ is irrational or $b > a$, and therefore
  $b >= 2$. If $a/b = 1/2$, the projection is a square, which has four sides. 
\end{proof}
